Oregon GWR smoke association with Cardiopulmonary Emergency Room visits (PartA)

### 2018-04-16 r_scripts/analysis/oregon_part_a.Rmd
oregon part A markdown file. GWR smoke association with sameday ER cardiopulmonary visits.

---
title: "Oregon: Zip-Level Wildfire Smoke and Health Outcomes (open in script)"
author: "Jingyang Liu"
date: "April 7, 2017"
output: html_document
---

@@

This document contains results for the association between wildfire smoke PM~2.5~ and cardiovascular (CVD) and respiratory health outcomes. This particular markdown document focuses on the zip code-level analysis. Herein, I compare different PM~2.5~ estimation methods described in further detail below.

### Time-Stratified Case-Crossover (referents same day of week within fire season May - September)

```{r include=FALSE}
library(tidyverse)
library(data.table)
library(survival)
library(htmlTable)
library(readr)
```

```{r ER data import, include = F, eval = FALSE}

### Limit to only ER visits
## pos = 23
var_list <- c('respiratory', 'asthma', 'pneumonia',  'acute_bronch', 'copd', 
              'cvd', 'isch_heart_dis', 'arrhythmia', 'heart_failure', 
              'cerbrovas_dis', 'myocardial_infarc', 'broken_arm')

@@

num_person <- data.frame(rbind(var_list, N))
```

```{r ER and urgent data import, include = F, eval = FALSE}

### Limit to ER visits and urgent care
## pos = 23 or pos = 20
var_list <- c('respiratory', 'asthma', 'pneumonia',  'acute_bronch', 'copd', 
              'cvd', 'isch_heart_dis', 'arrhythmia', 'heart_failure', 
              'cerbrovas_dis', 'myocardial_infarc', 'broken_arm')

  or_disease <- or_disease %>% 
    # filter the unknown gender
    filter(gender!= "U")

  # Create a permanent case-cross over dataset
  file_name <- paste('../data_new/casecrossover/ER_Urgent/or', i, 'may_to_sep_time_strat_casecross_both.csv', sep = "_")

num_person_both <- data.frame(rbind(var_list, N))
```

```{r urgent data import, include = F, eval = FALSE}

### Limit to ER visits and urgent care
## pos = 23 or pos = 20
var_list <- c('respiratory', 'asthma', 'pneumonia',  'acute_bronch', 'copd', 
              'cvd', 'isch_heart_dis', 'arrhythmia', 'heart_failure', 
              'cerbrovas_dis', 'myocardial_infarc', 'broken_arm')

  or_disease <- or_disease %>% 
    filter(pos=="20") %>%
    # filter the unknown gender
    filter(gender!= "U")

  # Create a permanent case-cross over dataset
  file_name <- paste('../data_new/casecrossover/Urgent/or', i, 'may_to_sep_time_strat_casecross_ur.csv', sep = "_")

num_person_ur <- data.frame(rbind(var_list, N))
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
## Analysis
path <- paste0("../data_new/casecrossover/ER_Urgent")

resp_casecross <- read_csv(paste(path, 
                                 "or_respiratory_may_to_sep_time_strat_casecross_both.csv", 
                                 sep = "/"))

# asthma
asthma_casecross <- read_csv(paste(path, 
                                   "or_asthma_may_to_sep_time_strat_casecross_both.csv", 
                                   sep = "/"))

# copd 
copd_casecross <- read_csv(paste(path, 
                                 "or_copd_may_to_sep_time_strat_casecross_both.csv", 
                                 sep = "/"))

# pneum or bronchitis
pneum_casecross <- read_csv(paste(path, 
                                  "or_pneumonia_may_to_sep_time_strat_casecross_both.csv",
                                  sep="/"))
 # acute bronchitis
acute_bronch_casecross <- read_csv(paste(path, 
                                         "or_acute_bronch_may_to_sep_time_strat_casecross_both.csv",
                                         sep = "/"))
# cvd
cvd_casecross <- read_csv(paste(path, 
                                "or_cvd_may_to_sep_time_strat_casecross_both.csv",
                                sep="/"))
# arrhythmia
arrhythmia_casecross <- read_csv(paste(path, 
                                       "or_arrhythmia_may_to_sep_time_strat_casecross_both.csv",
                                       sep="/"))
# cerebral vascular
cereb_vas_casecross <- read_csv(paste(path, 
                                      "or_cerbrovas_dis_may_to_sep_time_strat_casecross_both.csv",
                                      sep="/"))
# heart failure
hf_casecross <- read_csv(paste(path, 
                               "or_heart_failure_may_to_sep_time_strat_casecross_both.csv", 
                               sep="/"))
# ischemic heart disease
ihd_casecross <- read_csv(paste(path, 
                                "or_isch_heart_dis_may_to_sep_time_strat_casecross_both.csv",
                                sep="/"))
# myo infarc
mi_casecross <- read_csv(paste(path, 
                               "or_myocardial_infarc_may_to_sep_time_strat_casecross_both.csv", sep="/"))

# broken arm
broken_arm_casecross <- read_csv(paste(path, 
                                       "or_broken_arm_may_to_sep_time_strat_casecross_both.csv",
                                       sep="/"))

# read zip pm for counts by zip
smoke_path <- paste0("../data/Oregon_PM/")
zip_pm <- read_csv(paste0(smoke_path, "zip_pm_to_merge_with_acap.csv"))
```

In these comparisons, we examine various methods of smoke/PM~2.5~ estimations and associations with health outcomes using a time-stratified case-crossover study design. Health outcomes for a patient with a primary diagnosis of cardiopulmonary health outcomes and their date of admission  (index time) were identified. We then created counterfactual observations for each patient for the same day of the week for the entire wildfire season (MAy 1st to Sep 30th, 2013). We further limited our analyses to claims that were coded as emergency or urgent visits to eliminate bias from patients going in for elective/planned procedures. However, we found the filtered population is much larger than the population we chose in Washington, so we choose only the emegency room (hospital) visits and urgent care visits. For patients who have a index time closer to May 1, 2013 or closer to September 30, 2013, their referent observations before or after these dates will be excluded as I will not be able to assign estimates of PM~2.5~ to these referent observations. However, this is not a big issue for a time-stratified design as other referent values throughout the time period average out.

The **health outcomes** of interest are all respiratory, asthma, COPD, pneumonia, acute bronchitis, cardiovascular disease, heart failure, ischemic heart disease, and myocardial infarction. I also included broken arm, which I hypothesize to not be associated with wildfire smoke exposure and use as a check.

As for **exposure methods**, there are three main estimation methods for PM~2.5~: WRF-Chem Smoke (which substracts the WRF-Chem no fire emission from WRF-chem). For the WRF-Chem variable with the 'smoke' designator, this is WRF-Chem - WRF-Chem no fire. For Geo-Weighted Regression, and Kriging with 'smk' designator, I subtracted off the 'Background' estimates of smoke, which I believe are the monthly averages of PM~2.5~ for a given grid. However, the 'Background' is the mean of total period from May 1st to Sep 30th 2013 so I have to assign the mean for every day.

The **analytic method** is the conditional logistic regression model using the *survival* package in R. Each conditional logistic regression model accounts for the subject, and adjusts for temperature from the WRF-Chem model. The conditional logistic regression model \beta represents the change in risk of an event associated with a short-term unit increase in exposure, and can be calculated as an average difference between exposure at the index time and a weighted average of exposure at all times in the referent window. See Janes et al. 2005, **Statistics in Medicine**.

I've run a couple different scenarios for selecting referent periods, each of which has some paper that may justify the benefit of the design, relative to other methods.

## Population-Weighted PM~2.5~ Assigned by Zip Code

### Descriptives

For all of 2013, there were a total of millons of all claims records in the APAC dataset. During the May 1st to September 30th wildfire season in Oregon state, there were 47413 emergency visits for the specified cardiopulmonary codes. The number of persons (cases), as well as age-specific and sex-specific strate.

@@

# dataframe list
df_list <- list(resp_casecross, asthma_casecross, copd_casecross, 
  pneum_casecross, acute_bronch_casecross, cvd_casecross, arrhythmia_casecross,
  cereb_vas_casecross, hf_casecross, ihd_casecross, mi_casecross, 
  broken_arm_casecross)

outcome_list <- c('All Respiratory', 'Asthma', 'COPD', 
                  'Pneumonia', 'Acute Bronchitis', 'Cardiovascular Disease',
                  'Arrhythmia', 'Cerebrovascular Disease', 'Heart Failure',
                  'Ischemic Heart Disease', 'Myocardial Infarction',
                  'Broken Arm')

method_list <- c('WRF-Chem Smoke', 'Geo-Weighted Smoke', 'Kriging Smoke')

@@

  # dataframe for analysis creation
  # bind columns back together 
  df_analysis <- df_to_loop %>%
    filter(!is.na(wrf_smk_pm_zip)) %>% 
    # only look at outcomes
    filter(outcome == 1) %>%
    # add another row that makes sure there is a person <15 in the dataframe
    # tricking xtabs to produce a 0 cell for the outcome for age <15
    add_row(outcome = 0, age_ind = 0)

@@

@@

df_list <- list(resp_casecross, asthma_casecross, copd_casecross, 
                pneum_casecross, acute_bronch_casecross, cvd_casecross, arrhythmia_casecross,
                cereb_vas_casecross, hf_casecross, ihd_casecross, mi_casecross, 
                broken_arm_casecross)

outcome_list <- c('All Respiratory', 'Asthma', 'COPD', 
                  'Pneumonia', 'Acute Bronchitis', 'Cardiovascular Disease',
                  'Arrhythmia', 'Cerebrovascular Disease', 'Heart Failure',
                  'Ischemic Heart Disease', 'Myocardial Infarction',
                  'Broken Arm')

method_list <- c('WRF-Chem Smoke', 'Kriging Smoke', 'Geo-Weighted Smoke')

@@

  # extract covariates from dataframe
  covariates_df <- df_to_loop[, c(1:26, 71, 74, 76:85)]

  # extract pm values and divide by 10 and ordered
  which(colnames(df_to_loop)=="background_pm_zip") # code to find column numbers, 91
  pm_estimates_df <- df_to_loop[, c(88, 93, 92, 94)]/10  # create 10 unit increases

  # dataframe for analysis creation
  # bind columns back together 
  df_analysis <- cbind(covariates_df, pm_estimates_df) %>% 
    # remove missing pm values
    filter(!is.na(wrf_smk_pm_zip)) %>%

@@

  # empty matrix
  point_estimates <- matrix(nrow = 3, ncol = 9, byrow = T)

@@

  # second loop to run a model for each pm estimation method
  for(j in 39:41){

@@

    # conditional logistic regression model
    mod <- clogit(outcome ~ df_analysis[[j]] + wrf_temp_zip + strata(personkey), df_analysis)

@@

  }

@@

@@

tab <- htmlTable(txtRound(table_df, digits = 3, 1:3), 
                 caption = "Association between a 10 ug/m^3 in PM2.5 and Health Outcomes",
                 # row group by outcome
                 rgroup = outcome_list,
                 n.rgroup = c(rep(3, 12)), # 4 rows for each method for each outcome
                 # column headers
                 header = c("Method", "Obs.", "Events",
                            "OR&dagger;", "Lower", "Upper"),
                 # column spanner
                 cgroup = c("", "95% CI"), 
                 n.cgroup = c(4, 2),
                 padding.rgroup = "&nbsp;&nbsp;",
                 css.cell = "padding-left: 0.5em; padding-right: .5em;", # cell space
                 align = "llccccc", # column alignment,
                 tfoot="&dagger; Adjusted for CDC temperature, relatively humidity, wind speed, and precipitation; accounting for subject. Time-stratified: referent periods matched to events on same day of week within July to October fire season."
) # end table

@@

print(print_plot)
# save figure
ggsave("../plot_new/fig_both_er_ur.pdf", plot = print_plot, 
       width = 12, height = 8, units = "in")
```

#### Result Summary: Time-Stratified Case-Crossover Design (same day of week within the season)

A general interpretation is as follows:
A 10 $\mu$g/m^3^ increase in PM~2.5~ attributed to wildfire smoke was associated with a X% increase in the risk for a health event. Where X% is the odds ratio - 1 (e.g. OR of 1.20 - 1 = 20% increase). 
Note that the odds ratio approximates the relative risk in this case due to the rare outcome assumption in that the calcuation for an odds ratio is similar to risk ratio calcuation.

We see that all exposure methods are significantly associated with an increase in all respiratory outcomes and asthma. The WRF-Chem smoke method is associated with cerebrovascular disease as well. As for the other models, Kriging, Global Regression, and Geo-Weighted Regression, those are associated with COPD. I believe this is the most justifiable design approach that would average any extreme values of PM~2.5~ across the wildfire season for the vector of referent values for a particular subject.

#### Wildfire Smoke and Oregon All Payer All Claims Results
We found a  10 $\mu$g/m^3^ increase in PM~2.5~ was significantly associated with an 2.8% increase for respiratory oucomes  using method of WRF-Chem smoke (OR: 1.028 95% Confidence Interval [CI]: 1.001-1.055), and no significant association for kriging smoke method (OR: 1.018, 95% CI: 0.998-1.038), and also no significant association for GWR smoke method (OR: 1.021, 95% CI: 0.999-1.043)
A 10 $\mu$g/m^3^ increase in PM~2.5~ was significantly associated with an 9.0% increase for asthma outcomes using  WRF-Chem smoke method (OR: 1.090, 95% Confidence Interval [CI]: 1.014-1.171), and 7.9% increase for kriging smoke method (OR: 1.079, 95% CI: 1.036-1.123), and 8.5% increase for GWR smoke method (OR: 1.085, 95% CI: 1.038-1.134)

```{r lags with cdc met adj, warning = F, echo = F, results='asis', eval=FALSE, include = F} 
# note above, results = 'asis' needed to print htmlTables

# dataframe list
df_list <- list(resp_casecross, asthma_casecross, copd_casecross, pneum_casecross,
                acute_bronch_casecross, cvd_casecross, arrhythmia_casecross,
                cereb_vas_casecross, hf_casecross, ihd_casecross, mi_casecross, 
                broken_arm_casecross)

outcome_list <- c('All Respiratory', 'Asthma', 'COPD', 'Pneumonia', 
                  'Acute Bronchitis', 'Cardiovascular Disease', 'Arrhythmia', 
                  'Cerebrovascular Disease', 'Heart Failure',
                  'Ischemic Heart Disease', 'Myocardial Infarction', 'Broken Arm')

# looking only at wrf chem, kriging, and geo-weighted; taking out 'Global Smoke' method
method_list <- c('WRF-Chem Smoke', 'Kriging Smoke', 'Global Smoke', 'GWR Smoke')
method_list <- c('WRF-Chem Smoke', 'Kriging Smoke', 'Geo-Weighted Smoke')

# data wrangling ----
# Producing conditional logit model estimates loop 
for(i in 1:length(df_list)){
 i = 2

  # extract covariates from dataframe (Oregon)
  covariates_df <- df_to_loop[, c(1:26, 71, 74, 76:85)]

  # extract pm values and divide by 10 and ordered
  # which(colnames(df_to_loop)=="wrf_smk_pm_lag1_zip") # code to find column numbers (105)
  # which(colnames(df_to_loop)=="wrf_smk_pm_zip") # code to find column numbers (88)
  # which(colnames(df_to_loop)=="krig_smk_pm_lag1_zip") # code to find column numbers (130)
  # which(colnames(df_to_loop)=="krig_smk_pm_zip") # code to find column numbers (93)
  # which(colnames(df_to_loop)=="geo_smk_pm_lag1_zip") # code to find column numbers (125)
  # which(colnames(df_to_loop)=="geo_smk_pm_zip") # code to find column numbers (92)
  # which(colnames(df_to_loop)=="wrf_temp_lag1_zip") # code to find column numbers (135)
  # which(colnames(df_to_loop)=="wrf_temp_zip") # code to find column numbers (94)

  pm_lag_list <- list(df_to_loop[, c(88, 105:109)]/10, df_to_loop[, c(93, 130:134)]/10,
                  df_to_loop[, c(92, 125:129)]/10, df_to_loop[, c(94, 135:139)]/10)  # create 10 unit increases

# need to think if I need a methods section (probably not since this is it)

  # new loop for each element of the lag list 
  for(k in 1:4){
    k = 1
    # empty matrix (12 x 10 matrix)
    point_estimates <- matrix(nrow = 6, ncol = 10, byrow = T)

    colnames(point_estimates) <- c('outcome', 'pm_method', 'lag', 'n', 'n_events', 
                                   'odds_ratio', 'lower95', 'upper95', 'se', 'p_val')

    # fill in the outcome namedataframe before method loop
    point_estimates[, 1] <- outcome_name
    # repeat the sex category 4 times for each pm method
    point_estimates[, 2] <- method_list[k]

    pm_test <- pm_lag_list[[k]]

    # dataframe for analysis creation
    # bind columns back together 
    df_analysis <- cbind(covariates_df, pm_lag_list[[k]])
#glimpse(df_analysis)

  # loop to run a model for each lag 0-5
    for(j in 0:5){

      j = 0

      # finding locations of columns in dataframe
      # which(colnames(df_analysis)=="daily_meanWS")

      # set column number to evaluate for the pm lag method
      #pm_lag_colnum <- j + 31
      # set column number of cdc temp lag to adjust for
      #temp_lag_colnum <- j + 49
      # set col num for abs temp
      #ab_temp_lag_colnum <- j + 37
      # set col num for relative humidity
      #rh_lag_colnum <- j + 43
      # precip
      #pr_lag_colnum <- j + 55
      # wind speed
      #ws_lag_colnum <- j + 61
      # set row number to fill
      #row_n <- j+1

      # only run the model if the dataframe has observations
      if(nrow(df_analysis) != 0){
      # conditional logistic regression model
     #  mod <- clogit(outcome ~  df_analysis[[pm_lag_colnum]] + df_analysis[[temp_lag_colnum]] + 
     #                df_analysis[[rh_lag_colnum]] + df_analysis[[pr_lag_colnum]] + 
     #                df_analysis[[ws_lag_colnum]] +strata(personkey), df_analysis)

        mod <- clogit(outcome ~ df_analysis[[j]] + wrf_temp_zip + strata(personkey), df_analysis)

      # populate matrix
      point_estimates[row_n, 3] <- j
      point_estimates[row_n, 4] <- mod$n
      point_estimates[row_n, 5] <- mod$nevent
      # odds ratio
      point_estimates[row_n, 6] <- round(exp(summary(mod)$coefficient[1,1]), 3)
      # 95% lower bound
      point_estimates[row_n, 7] <- round(exp((summary(mod)$coefficient[1,1]) -
                                        1.96*(summary(mod)$coefficient[1,3])), 3)
      # 95% upper bound
      point_estimates[row_n, 8] <- round(exp((summary(mod)$coefficient[1,1]) +
                                        1.96*(summary(mod)$coefficient[1,3])), 3)
      # standard error
      point_estimates[row_n, 9] <- round(summary(mod)$coefficient[1,3], 4)
      # p val
      point_estimates[row_n, 10] <- round(summary(mod)$coefficient[1,5], 4)

    } # end methods loop

  # bind rows of sex category estimates together
 lag_est_df <- bind_rows(datalist1)

 # populate second dataframe list
 datalist2[[i]] <- lag_est_df

#glimpse(combined_point_est_df)
# subset to just geo-smoke estimates for now to reduce number of rows
wrf_geo_lag <- combined_point_est_df %>% 
  # filter columns to just geo-smk
  filter(pm_method == "GWR Smoke" |
         pm_method == "WRF-Chem Smoke") %>% 
  # subset columns I want to put in to the table
  select(2:3, 5:8) %>% 
  mutate(odds_ratio = ifelse(odds_ratio == 99, "--", odds_ratio),
         lower95 = ifelse(lower95 == 99, "--", lower95),
         upper95 = ifelse(upper95 == 99, "--", upper95))

  tab <- htmlTable(txtRound(wrf_geo_lag, digits = 3, 1:3), 
           caption = "Association between a 10 ug/m^3 in PM2.5 geo smoke and health outcomes by lagged days",
           # row group by outcome
           rgroup = outcome_list,
           n.rgroup = c(rep(12, 12)), # 6 rows for each lag period for each outcome
           # column headers
           header = c("Est Method", "Lagged Days", "Events",
                      "OR&dagger;", "Lower", "Upper"),
           # column spanner
           cgroup = c("", "95% CI"), 
           n.cgroup = c(4, 2),
           padding.rgroup = "&nbsp;&nbsp;",
           css.cell = "padding-left: 0.5em; padding-right: .5em;", # cell space
           align = "lccccc", # column alignment,
           tfoot="&dagger; Adjusted for lagged CDC temperature, windspeed, relative humidity, and precipitation, accounting for subject."
            ) # end table

# ggplot of odds ratios, facet wrapped by outcomes -----
# convert variables from character to either numeric or factor
# factor preserves the order of the variable
wrf_geo_lag_plot <- combined_point_est_df %>% 
  # filter columns to just geo-smk
  filter(pm_method == "GWR Smoke" |
         pm_method == "WRF-Chem Smoke") %>% 
  # subset columns I want to put in to the table
  select(1:3, 5:8)

# change characters to numeric and factor  
wrf_geo_lag_plot$outcome <- factor(wrf_geo_lag_plot$outcome,
                              levels = unique(wrf_geo_lag_plot$outcome))

wrf_geo_lag_plot$pm_method <- factor(wrf_geo_lag_plot$pm_method,
                                levels = unique(wrf_geo_lag_plot$pm_method))

wrf_geo_lag_plot$lag <- factor(wrf_geo_lag_plot$lag,
                                levels = unique(wrf_geo_lag_plot$lag))

wrf_geo_lag_plot$n_events <- as.numeric(wrf_geo_lag_plot$n_events)
wrf_geo_lag_plot$odds_ratio <- as.numeric(wrf_geo_lag_plot$odds_ratio)
wrf_geo_lag_plot$lower95 <- as.numeric(wrf_geo_lag_plot$lower95)
wrf_geo_lag_plot$upper95 <- as.numeric(wrf_geo_lag_plot$upper95)

## ggplot
  print_plot <- ggplot(wrf_geo_lag_plot,
                  aes(x = lag, y = odds_ratio, color = pm_method), na.rm = T) +
    geom_point(position = position_dodge(width = 0.5), na.rm = T) + 
    geom_errorbar(aes(ymin=lower95, ymax=upper95), 
                  position = position_dodge(width = 0.5), width = 0.2, na.rm =T) +
    # custom color 
    scale_color_manual(name = "Smoke-Estimation Method", 
                       values = c("red", "#32115C"),
                       guide = guide_legend(title.position = "top", title.hjust = 0.5)) +
    facet_wrap(~outcome, nrow = 3, scales = "free") +
    geom_hline(yintercept = 1, linetype=2) +
    #ggtitle('Association Between PM2.5 from \n Wildfire Smoke on Hospitalizations') +
    ylab(expression(paste("Odds Ratio for 10µg/m"^3, " Increase in PM"[2.5]))) +
    #ylim(0.7, 1.4) +
    xlab('Lag Days') +
    # plot theme
    theme(panel.background = element_rect(fill = 'white', colour = 'black'),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    # strip element
    strip.background = element_rect(colour=NA, fill=NA),
    panel.border = element_rect(fill = NA, color = "black"),
    # facet text size
    strip.text = element_text(size = 10),
    # axis element
    #axis.text.x = element_blank(),
    #axis.title.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.y = element_text(angle = 90),
    # legend elements
    legend.position = "bottom")
    #legend.text = element_text(size = 8))

  print(print_plot)

  # save figure
  ggsave("../plot_new/fig4.pdf", plot = print_plot, 
       width = 12, height = 8, units = "in")

@@

Following table and figure look at some outcomes stratified by age category. Note that kids <=15 likely won't have many outcomes like COPD or MI. I have restricted the dataset to just the geo smoke method to reduce the number of rows/points on the tables and figures to make it easier to read for now. Code can easily be changed to have all PM~2.5~ methods.

@@

# dataframe list
df_list <- list(resp_casecross, asthma_casecross, copd_casecross, pneum_casecross,
                acute_bronch_casecross, cvd_casecross, arrhythmia_casecross,
                cereb_vas_casecross, hf_casecross, ihd_casecross, mi_casecross, 
                broken_arm_casecross)

outcome_list <- c('All Respiratory', 'Asthma', 'COPD', 'Pneumonia', 
                  'Acute Bronchitis', 'Cardiovascular Disease', 'Arrhythmia', 
                  'Cerebrovascular Disease', 'Heart Failure',
                  'Ischemic Heart Disease', 'Myocardial Infarction', 'Broken Arm')

# looking only at wrf chem, kriging, and geo-weighted; taking out 'Global Smoke' method
method_list <- c('WRF-Chem Smoke', 'Kriging Smoke', 'GWR Smoke')

@@

var_list <- c('respiratory', 'asthma', 'pneumonia',  'acute_bronch', 'copd', 
              'cvd', 'isch_heart_dis', 'arrhythmia', 'heart_failure', 
              'cerbrovas_dis', 'myocardial_infarc', 'broken_arm')

@@

  # extract covariates from dataframe
  covariates_df <- df_to_loop[, c(1:26, 71, 74, 76:85)]

  # extract pm values and divide by 10 and ordered
  #which(colnames(df_to_loop)=="global_smk_pm_zip") # code to find column numbers
  pm_estimates_df <- df_to_loop[, c(88, 93, 92, 94)]/10  # create 10 unit increases

@@

    } # end methods loop

>>>>

---
title: "Oregon GWR smoke association with Cardiopulmonary Emergency Room visits (PartA)"
author: "Jingyang Liu"
date: "April 9, 2018"
output: html_document
---

@@

This project researches on for the association between GWR smoke PM~2.5~ and cardiopulmonary (cardiovascular (CVD) and respiratory health outcomes) diseases for Oregon State in wildfire season (May to September) in 2013. 
This markdown document focuses on the zip code-level analysis. 
The analysis includes total analysis, age strata and sex strata.

In these research, we mainly examine GWR smoke/PM~2.5~ estimations and associations with health outcomes using a time-stratified case-crossover study design. Health outcomes for a patient with a primary diagnosis of cardiopulmonary health outcomes and their date of admission  (index time) were identified. We then created counterfactual observations for each patient for the same day of the week for the entire wildfire season (MAy 1st to Sep 30th, 2013). We further limited our analyses to claims that were coded as emergency or urgent visits to eliminate bias from patients going in for elective/planned procedures. However, we found the filtered population is much larger than the population we chose in Washington, so we choose only the emegency room (hospital) visits and urgent care visits. For patients who have a index time closer to May 1, 2013 or closer to September 30, 2013, their referent observations before or after these dates will be excluded as I will not be able to assign estimates of PM~2.5~ to these referent observations. However, this is not a big issue for a time-stratified design as other referent values throughout the time period average out.

The **cardiopulmonary health outcomes** are all respiratory, asthma, COPD, cardiovascular disease, heart failure, and ischemic heart disease.

As for **exposure methods**, we use Geo-Weighted Regression and subtracted off the 'Background' estimates of smoke, which I believe are the monthly averages of PM~2.5~ for a given grid. However, the 'Background' is the mean of total period from May 1st to Sep 30th 2013 so I have to assign the mean for every day. Then use the 10 units of the smoke for analysis.

The **analytic method** is the conditional logistic regression model using the *survival* package in R. This regression is controlling the confounders. Each conditional logistic regression model accounts for the subject, and adjusts for temperature from the WRF-Chem model. The conditional logistic regression model \beta represents the change in risk of an event associated with a short-term 10 units' increase in exposure, and can be calculated as an average difference between exposure at the index time and a weighted average of exposure at all times.

**Library:**  
tidyverse is for data cleaning, also includes ggplot and readr.  
data.table is for quick reading files.  
htmlTable is for writing table.  
survival is for running conditional logistic regression model.

```{r library, include=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(data.table)
library(survival)
library(htmlTable)
```

**Time-Stratified Case-Crossover Data Set (same day of week within wildfire season May - September)**

The data set is from the case crossover data set, which is made before. To save the running time, the data set is wildfire season and people's first visit for same date. Then filter the Emergency postal of service and non-missing gender visits.

```{r ER data making, include = F, eval = FALSE}
### Limit to only ER visits
## pos = 23
var_list <- c('respiratory', 'asthma', 'pneumonia',  'acute_bronch', 'copd', 
              'cvd', 'isch_heart_dis', 'arrhythmia', 'heart_failure', 
              'cerbrovas_dis', 'myocardial_infarc')

@@

# check the number of people in each disease
num_person <- data.frame(rbind(var_list, N))
```

Then load the Emergency data set.

```{r load ER data set and zip smoke, echo=FALSE, message=FALSE, warning=FALSE}
## Analysis
path <- paste0("../data_new/casecrossover/ER/")

resp_casecross <- read_csv(paste(path, 
                                 "or_respiratory_may_to_sep_time_strat_casecross_er.csv", 
                                 sep = "/"))

# asthma
asthma_casecross <- read_csv(paste(path, 
                                   "or_asthma_may_to_sep_time_strat_casecross_er.csv", 
                                   sep = "/"))

# copd 
copd_casecross <- read_csv(paste(path, 
                                 "or_copd_may_to_sep_time_strat_casecross_er.csv", 
                                 sep = "/"))

# pneum or bronchitis
pneum_casecross <- read_csv(paste(path, 
                                  "or_pneumonia_may_to_sep_time_strat_casecross_er.csv",
                                  sep="/"))
 # acute bronchitis
acute_bronch_casecross <- read_csv(paste(path, 
                                         "or_acute_bronch_may_to_sep_time_strat_casecross_er.csv",
                                         sep = "/"))
# cvd
cvd_casecross <- read_csv(paste(path, 
                                "or_cvd_may_to_sep_time_strat_casecross_er.csv",
                                sep="/"))
# arrhythmia
arrhythmia_casecross <- read_csv(paste(path, 
                                       "or_arrhythmia_may_to_sep_time_strat_casecross_er.csv",
                                       sep="/"))
# cerebral vascular
cereb_vas_casecross <- read_csv(paste(path, 
                                      "or_cerbrovas_dis_may_to_sep_time_strat_casecross_er.csv",
                                      sep="/"))
# heart failure
hf_casecross <- read_csv(paste(path, 
                               "or_heart_failure_may_to_sep_time_strat_casecross_er.csv", 
                               sep="/"))
# ischemic heart disease
ihd_casecross <- read_csv(paste(path, 
                                "or_isch_heart_dis_may_to_sep_time_strat_casecross_er.csv",
                                sep="/"))
# myo infarc
mi_casecross <- read_csv(paste(path, 
                               "or_myocardial_infarc_may_to_sep_time_strat_casecross_er.csv", sep="/"))

### Descriptive Table

This table shows number of cardiopulmonary visits for each outcome observed from May 1st to September 30st in 2013. Sex and Age strata for each outcome are also shown in this table. Female and Male show similar proportion in each outcome. As for age groups, adult group shows most proportion in asthma outcome than other age groups, and also it is the more likely disease adults tend to have. Older people tend to get cardiovascular disease while children tend to get respiratory diseases, which makes sense.

*Missing smoke value data is already removed for table and next analysis, although only a few.*

@@

# dataframe list
df_list <- list(resp_casecross, asthma_casecross, copd_casecross, 
  pneum_casecross, acute_bronch_casecross, cvd_casecross, arrhythmia_casecross,
  cereb_vas_casecross, hf_casecross, ihd_casecross, mi_casecross)

outcome_list <- c('All Respiratory', 'Asthma', 'COPD', 
                  'Pneumonia', 'Acute Bronchitis', 'Cardiovascular Disease',
                  'Arrhythmia', 'Cerebrovascular Disease', 'Heart Failure',
                  'Ischemic Heart Disease', 'Myocardial Infarction')

method_list <- c('Geo-Weighted Smoke')

@@

  # dataframe for analysis creation
  # bind columns back together 
  df_analysis <- df_to_loop %>%
    filter(!is.na(geo_smk_pm_zip)) %>% 
    # only look at outcomes
    filter(outcome == 1) %>%
    # add another row that makes sure there is a person <15 in the dataframe
    # tricking xtabs to produce a 0 cell for the outcome for age <15
    add_row(outcome = 0, age_ind = 0)

@@

### Total Outcome Analysis

The following table shows the total outcome without strata.

@@

df_list <- list(resp_casecross, asthma_casecross, copd_casecross, 
                pneum_casecross, acute_bronch_casecross, cvd_casecross, arrhythmia_casecross,
                cereb_vas_casecross, hf_casecross, ihd_casecross, mi_casecross)

outcome_list <- c('All Respiratory', 'Asthma', 'COPD', 
                  'Pneumonia', 'Acute Bronchitis', 'Cardiovascular Disease',
                  'Arrhythmia', 'Cerebrovascular Disease', 'Heart Failure',
                  'Ischemic Heart Disease', 'Myocardial Infarction')

method_list <- c('Geo-Weighted Smoke')

@@

  # extract covariates from dataframe
  covariates_df <- df_to_loop[, c(1:26, 71, 74, 76:85, 92, 94)] %>%
    mutate(geo_smk_pm_zip10 = geo_smk_pm_zip/10)

  # dataframe for analysis creation
  # bind columns back together 
  df_analysis <- covariates_df %>% 
    # remove missing pm values
    filter(!is.na(geo_smk_pm_zip10)) %>%

@@

  # empty matrix
  point_estimates <- matrix(nrow = 1, ncol = 9, byrow = T)

@@

  # second loop to run a model for each pm estimation method

  j <- 39

@@

    # conditional logistic regression model
    mod <- clogit(outcome ~ geo_smk_pm_zip10 + wrf_temp_zip + strata(personkey), df_analysis)

@@

@@

# write_path <- paste0('../data_new/casecrossover/geo_health_or.csv')
# write_csv(combined_point_est_df, write_path)

@@

tab <- htmlTable(txtRound(table_df, digits = 3, 1:3), 
                 caption = "Association between a 10 ug/m^3 in PM2.5 and Health Outcomes",
                 # row group by outcome
                 rgroup = outcome_list,
                 n.rgroup = c(rep(1, 11)), # 4 rows for each method for each outcome
                 # column headers
                 header = c("Method", "Obs.", "Events",
                            "OR&dagger;", "Lower", "Upper"),
                 # column spanner
                 cgroup = c("", "95% CI"), 
                 n.cgroup = c(4, 2),
                 padding.rgroup = "&nbsp;&nbsp;",
                 css.cell = "padding-left: 0.5em; padding-right: .5em;", # cell space
                 align = "llccccc", # column alignment,
                 tfoot="&dagger; Adjusted for CDC temperature, relatively humidity, wind speed, and precipitation; accounting for subject. Time-stratified: referent periods matched to events on same day of week within May to Sep fire season."
) # end table

@@

# print(print_plot)
# save figure
# ggsave("../plot_new/fig_both_er_ur.pdf", plot = print_plot, 
#        width = 12, height = 8, units = "in")

write_path <- paste0("./all_strata_result_table.csv")
write_csv(combined_point_est_df, write_path)

@@

Following table and figure shows some outcomes stratified by age category. Note that kids <=15 likely does not have many outcomes like COPD or MI. I have restricted the dataset to just the geo smoke method to reduce the number of rows/points on the tables and figures to make it easier to read for now.

@@

# dataframe list
df_list <- list(resp_casecross, asthma_casecross, copd_casecross, pneum_casecross,
                acute_bronch_casecross, cvd_casecross, arrhythmia_casecross,
                cereb_vas_casecross, hf_casecross, ihd_casecross, mi_casecross)

outcome_list <- c('All Respiratory', 'Asthma', 'COPD', 'Pneumonia', 
                  'Acute Bronchitis', 'Cardiovascular Disease', 'Arrhythmia', 
                  'Cerebrovascular Disease', 'Heart Failure',
                  'Ischemic Heart Disease', 'Myocardial Infarction')

# looking only at wrf chem, kriging, and geo-weighted; taking out 'Global Smoke' method
method_list <- c('GWR Smoke')

@@

var_list <- c('respiratory', 'asthma', 'pneumonia',  'acute_bronch', 'copd', 
              'cvd', 'isch_heart_dis', 'arrhythmia', 'heart_failure', 
              'cerbrovas_dis', 'myocardial_infarc')

@@

  # extract covariates from dataframe
  covariates_df <- df_to_loop[, c(1:26, 71, 74, 76:85, 92, 94)] %>%
    mutate(geo_smk_pm_zip10 = geo_smk_pm_zip/10)

@@
